---
title: "NYC Payroll Analysis: Identifying Potential Savings"
author: "Dev Lala"
date: "`r Sys.Date()`"
format: html
editor: visual
execute:
  engine: knitr
  freeze: true
  echo: true  
  warning: false  
  message: false  
  error: true 
---

### **Commission to Analyze Taxpayer Spending (CATS)**

#### **Acquiring and Preparing Payroll Data**

::: details
### Code

```{r}
library(tidyverse) 
library(stringr) 
library(readr)

if (!file.exists("local_data/nyc_payroll_export.csv")) {
  
  dir.create("local_data", showWarnings = FALSE, recursive = TRUE)
  
  ENDPOINT <- "https://data.cityofnewyork.us/resource/k397-673e.json"
  
  packages <- c("httr2", "jsonlite", "dplyr", "readr")
  
  for (pkg in packages) {
    if (!require(pkg, character.only = TRUE)) install.packages(pkg, dependencies = TRUE)
    library(pkg, character.only = TRUE)
  }
  
  BATCH_SIZE <- 50000
  OFFSET <- 0
  END_OF_EXPORT <- FALSE
  ALL_DATA <- list()
  
  while (!END_OF_EXPORT) {
    cat("Requesting items", OFFSET, "to", BATCH_SIZE + OFFSET, "\n")
    
    req <- request(ENDPOINT) |>
      req_url_query(`$limit`  = BATCH_SIZE, 
                    `$offset` = OFFSET)
    
    resp <- req_perform(req)
    batch_data <- fromJSON(resp_body_string(resp))
    ALL_DATA <- append(ALL_DATA, list(batch_data))
    
    if (NROW(batch_data) != BATCH_SIZE) {
      END_OF_EXPORT <- TRUE
      cat("End of Data Export Reached\n")
    } else {
      OFFSET <- OFFSET + BATCH_SIZE
    }
  }
  
  ALL_DATA <- bind_rows(ALL_DATA)
  
  cat("Data export complete:", NROW(ALL_DATA), "rows and", NCOL(ALL_DATA), "columns.\n")
  write_csv(ALL_DATA, "local_data/nyc_payroll_export.csv")
  
} else {
  cat("File already exists. Skipping download.\n")
}
payroll_data <- read_csv("local_data/nyc_payroll_export.csv")

glimpse(payroll_data) 
head(payroll_data)
```
:::

::: details
### Code

```{r}
payroll_cleaned <- payroll_data %>%
  mutate(
    agency_name = str_to_title(agency_name),
    last_name = str_to_title(last_name),
    first_name = str_to_title(first_name),
    work_location_borough = str_to_title(work_location_borough),
    title_description = str_to_title(title_description),
    leave_status_as_of_june_30 = str_to_title(leave_status_as_of_june_30)
  )

glimpse(payroll_cleaned)
head(payroll_cleaned)
```
:::

**Fixing Boroughs names**

::: details
### Code

```{r}
payroll_cleaned <- payroll_cleaned %>%
  mutate(
    work_location_borough = case_when(
      work_location_borough == "Richmond" ~ "Staten Island",
      work_location_borough %in% c("Orange", "Westchester", "Sullivan", "Nassau", 
        "Delaware", "Ulster", "Washington Dc", "Putnam", 
        "Schoharie", "Dutchess", "Greene", "Albany") ~ "Other",
      TRUE ~ work_location_borough
    )
  )

unique(payroll_cleaned$work_location_borough)
write_csv(payroll_cleaned, "local_data/nyc_payroll_cleaned.csv")
```
:::

::: details
### Code

```{r}
library(dplyr) 
library(DT)    
library(scales)
eric_adams_data <- payroll_cleaned %>%
  filter(first_name == "Eric" & last_name == "Adams") %>%
  select(fiscal_year, title_description, agency_name, regular_gross_paid) %>%
  rename(
    `Fiscal Year` = fiscal_year,
    `Position` = title_description,
    `Agency` = agency_name,
    `Total Salary` = regular_gross_paid
  ) %>%
  arrange(`Fiscal Year`)

```
:::

::: details
### Code

```{r}
eric_adams_data <- eric_adams_data %>%
  mutate(`Total Salary` = dollar(`Total Salary`))

datatable( eric_adams_data, options = list( searching = FALSE,paging = FALSE,info = FALSE) )
```
:::

**Calculating total compensation**

::: details
### Code

```{r}
payroll_cleaned <- payroll_cleaned %>%
  mutate(
    total_compensation = case_when(
      pay_basis == "per Annum" ~ base_salary,  
      pay_basis == "per Hour" ~ (regular_hours * base_salary) + (ot_hours * base_salary * 1.5),  # Hourly Employees
      pay_basis == "per Day" & title_description == "Adjunct Lecturer" ~ (regular_hours / 1.5) * base_salary,  # Adjunct Lecturers
      pay_basis == "per Day" ~ (regular_hours / 7.5) * base_salary,  
      TRUE ~ NA_real_  
    )
  )

glimpse(payroll_cleaned)
head(payroll_cleaned %>% select(first_name, last_name, pay_basis, base_salary, regular_hours, ot_hours, total_compensation))
```
:::

**Filtering Eric Adam's records**

::: details
### Code

```{r}
eric_adams_data <- payroll_cleaned %>%
  filter(first_name == "Eric", last_name == "Adams") %>%
  select(fiscal_year, title_description, agency_name, total_compensation) %>%
  arrange(fiscal_year)

head(eric_adams_data)
```
:::

::: details
### Code

```{r}
eric_adams_data <- payroll_cleaned %>%
  filter(first_name == "Eric", last_name == "Adams") %>%
  select(fiscal_year, title_description, agency_name, total_compensation) %>%
  arrange(fiscal_year)

library(scales)
eric_adams_data <- eric_adams_data %>%
  mutate(total_compensation = dollar(total_compensation))

library(DT)
datatable(
  eric_adams_data,
  options = list(searching = FALSE, paging = FALSE, info = FALSE))
```
:::

**Task4 Q1 - Finding the Job title with highest base pay rate**

::: details
### Code

```{r}
highest_paid_job <- payroll_cleaned %>% 
  filter(!is.na(base_salary) & base_salary > 0) %>%
  group_by(title_description) %>% 
  summarise(max_base_salary = max(base_salary, na.rm = TRUE))%>%
  arrange(desc(max_base_salary)) %>%
  slice(1)

highest_paid_job
```
:::

**Task4 Q2 - Finding the Individual with the Highest Total Payroll**

::: details
### Code

```{r}
highest_paid_individual <- payroll_cleaned %>%
  filter(!is.na(total_compensation) & total_compensation > 0) %>%  
  select(fiscal_year, first_name, last_name, total_compensation) %>%
  arrange(desc(total_compensation)) %>%
  slice(1)  

highest_paid_individual
```
:::

**Task4 Q3 - Finding the Individual Who Worked the Most Overtime Hours**

::: details
### Code

```{r}
most_overtime_individual <- payroll_cleaned %>% 
  filter(!is.na(ot_hours) & ot_hours > 0) %>%
  select(first_name, last_name, fiscal_year, ot_hours) %>%
  arrange(desc(ot_hours)) %>% 
  slice(1)

most_overtime_individual
```
:::

**Task4 Q4 -Finding the Agency with the Highest Average Total Annual Payroll**

::: details
### Code

```{r}
highest_avg_pay_agency <- payroll_cleaned %>%
  filter(!is.na(total_compensation)) %>%
  group_by(agency_name) %>%
  summarise(avg_total_pay = mean(total_compensation, na.rm = TRUE)) %>%
  arrange(desc(avg_total_pay)) %>%
  slice(1)  

highest_avg_pay_agency
```
:::

**Task 4 Q5 - Finding agency with the most employees on payroll in each year**

::: details
### Code

```{r}
most_employees_per_year <- payroll_cleaned %>% 
  group_by(fiscal_year, agency_name) %>% 
  summarise(employee_count = n(), .groups = "drop") %>% 
  arrange(fiscal_year, desc(employee_count)) %>% 
  group_by(fiscal_year) %>% 
  slice(1)

most_employees_per_year
```
:::

**Task4 Q6 - Finding agency with the highest overtime usage compared to regular hours**

::: details
### Code

```{r}
overtime_usage <- payroll_cleaned %>%
  group_by(agency_name) %>%
  summarise(
    total_regular_hours = sum(regular_hours, na.rm = TRUE),
    total_ot_hours = sum(ot_hours, na.rm = TRUE),
    ot_ratio = total_ot_hours / total_regular_hours  
  ) %>%
  arrange(desc(ot_ratio)) %>%  
  slice(1)  

overtime_usage
```
:::

**Task4 Q7 - Finding the average salary of employees who work outside the five boroughs**

::: details
### Code

```{r}
avg_salary_outside_nyc <- payroll_cleaned %>% 
  filter(!work_location_borough %in% 
           c("Manhattan", "Brooklyn", "Queens", "Bronx", "Staten Island")) %>% 
  summarise(average_salary = mean(base_salary, na.rm = TRUE))

avg_salary_outside_nyc
```
:::

**Task4 Q8 - Calculating city’s aggregate payroll grown over the past 10 years**

::: details
### Code

```{r}
payroll_growth <- payroll_cleaned %>%
  group_by(fiscal_year) %>%
  summarise(total_payroll = sum(total_compensation, na.rm = TRUE)) %>%
  arrange(fiscal_year)  


payroll_growth
```
:::

**Calculating Mayor's Salary for Each Fiscal Year**

::: details
### Code

```{r}
mayor_salary <- payroll_cleaned %>% 
  filter(title_description == "Mayor") %>% 
  select(fiscal_year, total_compensation) %>% 
  distinct()

mayor_salary
```
:::

**Finding employees who earn more than mayor**

::: details
### Code

```{r}
highest_paid_individual <- payroll_cleaned %>%
  filter(total_compensation == max(total_compensation, na.rm = TRUE)) %>%
  select(first_name, last_name, fiscal_year, title_description, agency_name, total_compensation)

highest_paid_individual
```
:::

**Including agency_name column in highest_paid_individual dataset**

::: details
### Code

```{r}
highest_paid_individual <- payroll_cleaned %>% 
  filter(total_compensation == max(total_compensation, na.rm = TRUE)) %>% 
  select(first_name, last_name, fiscal_year, title_description, agency_name, total_compensation)
```
:::

**Comparing high earning employees with mayor's salary**

::: details
### Code

```{r}
capped_salaries <- highest_paid_individual %>%
  left_join(mayor_salary, by = "fiscal_year", suffix = c("_employee", "_mayor")) %>%  
  
  mutate(
    capped_compensation = pmin(total_compensation_employee, total_compensation_mayor),  
    savings = total_compensation_employee - capped_compensation  
  )

#Calculating total savings among all affected employees
total_savings <- sum(capped_salaries$savings, na.rm = TRUE)


list(
  employees_affected = capped_salaries,  
  total_amount_saved = total_savings  
)
```
:::

::: details
### Code

```{r}
affected_agencies <- highest_paid_individual %>% 
  count(agency_name, sort = TRUE) %>% 
  head(10)

affected_job_titles <- highest_paid_individual %>% 
  count(title_description, sort = TRUE) %>% 
  head(10)

list(affected_agencies, affected_job_titles)
```
:::

::: details
### Code

```{r}
overtime_usage <- payroll_cleaned %>%
  group_by(agency_name, title_description) %>%
  summarise(total_overtime_hours = sum(ot_hours, na.rm = TRUE), .groups = "drop") %>%
  filter(total_overtime_hours > 0) 


overtime_usage <- overtime_usage %>%
  mutate(full_time_replacements = total_overtime_hours / 2000)


savings_analysis <- payroll_cleaned %>%
  group_by(agency_name, title_description) %>%
  summarise(
    avg_base_salary = mean(base_salary, na.rm = TRUE),
    total_overtime_hours = sum(ot_hours, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(total_overtime_hours > 0) %>%
  mutate(
    overtime_cost = total_overtime_hours * avg_base_salary * 1.5,  
    regular_cost = (total_overtime_hours / 2000) * avg_base_salary * 2000,  
    savings = overtime_cost - regular_cost  
  )


top_savings_agencies <- savings_analysis %>%
  group_by(agency_name) %>%
  summarise(total_savings = sum(savings, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_savings)) %>%
  head(10)  

list(
  overtime_usage = head(overtime_usage, 10),
  savings_analysis = head(savings_analysis, 10),
  top_savings_agencies = top_savings_agencies
)
```
:::

Summary

Our analysis of city payroll data reveals that high overtime costs, especially in the Police Department and Fire Department, make up a significant portion of overall expenses. Shifting overtime hours to full-time positions could lead to major savings, with the Police Department alone potentially saving around \$5.44 trillion. Other agencies, such as the Department of Corrections, NYC Housing Authority, and HRA, also rely heavily on overtime, which could be reduced through smarter hiring strategies.

**Recommendations**

Some recommendations are :

1.  Strategic Workforce expansion

2.  Overtime Management and Policy reform

3.  Comprehensive cost-benefit assessment

4.  Ongoing payroll and efficiency monitoring

    ::: details
    ### Code

    ```{r}
    overtime_by_job <- payroll_cleaned %>%
      group_by(agency_name, title_description) %>%
      summarise(total_ot_hours = sum(ot_hours, na.rm = TRUE), .groups = "drop") %>%
      arrange(desc(total_ot_hours))

    head(overtime_by_job, 10)
    ```
    :::

::: details
### Code

```{r}
overtime_replacement <- overtime_by_job %>% 
  mutate(full_time_needed = (total_ot_hours / 2080) * 1.5)

head(overtime_replacement, 10)
```
:::

::: details
### Code

```{r}
overtime_savings <- payroll_cleaned %>%
  group_by(agency_name, title_description) %>%
  summarise(
    total_ot_hours = sum(ot_hours, na.rm = TRUE),
    avg_hourly_wage = mean(base_salary / 2080, na.rm = TRUE),  
    .groups = "drop"
  ) %>%
  mutate(
    ot_cost = total_ot_hours * avg_hourly_wage * 1.5,  
    reg_cost = total_ot_hours * avg_hourly_wage,  
    savings = ot_cost - reg_cost  
  ) %>%
  arrange(desc(savings))


head(overtime_savings, 10)
```
:::

**Calculating savings for each agency**

::: details
### Code

```{r}
agency_savings <- overtime_savings %>% 
  group_by(agency_name) %>% 
  summarise(total_savings = sum(savings, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_savings))

head(agency_savings, 10)
```
:::

**Calculating estimated savings**

::: details
### Code

```{r}
total_savings <- sum(overtime_savings$savings, na.rm = TRUE)

total_savings
```
:::

## **New Policy recommendation**

### **Replacing Some Full-Time Roles with Part-Time Employees**

**Identify job titles that could be converted to part-time roles**

::: details
### Code

```{r}
part_time_eligible_roles <- payroll_cleaned %>% 
  filter(regular_hours == 2080) %>%
  group_by(title_description, agency_name) %>% 
  summarise( num_employees = n(), avg_salary_weekly = mean(base_salary / 52, na.rm = TRUE),.groups = "drop" ) %>% 
  mutate( full_time_cost = num_employees * 1200,part_time_cost = num_employees * 500 * 2,savings = full_time_cost - part_time_cost ) %>% 
  arrange(desc(savings))

head(part_time_eligible_roles, 10)

total_savings <- sum(part_time_eligible_roles$savings, na.rm = TRUE)

list( top_roles_for_conversion = head(part_time_eligible_roles, 10), total_potential_savings = total_savings )
```
:::

**Calculate part-time salary and cost savings**

::: details
### Code

```{r}
full_time_roles <- payroll_cleaned %>%
  filter(pay_basis == "per Annum") %>%  
  group_by(title_description, agency_name) %>%
  summarise(
    avg_salary = mean(base_salary, na.rm = TRUE),  
    num_employees = n(),
    .groups = "drop"
  )

part_time_conversion <- full_time_roles %>%
  mutate(
    part_time_salary = avg_salary * 0.6,  
    total_full_time_cost = avg_salary * num_employees,
    total_part_time_cost = part_time_salary * num_employees,
    savings = total_full_time_cost - total_part_time_cost
  )

head(part_time_conversion, 10)
```
:::

## **Summary of the New Policy**

This policy examines potential savings from replacing some full-time positions with part-time employees. Assuming part-time workers clock 20 hours per week while full-timers work 40, with wages adjusted accordingly, the analysis highlights significant cost reductions. The findings suggest that shifting select roles to part-time could cut costs while maintaining efficiency. The greatest savings come from jobs with high employee counts and lower skill requirements, where part-time coverage is feasible.

## **Recommendations**

-   Focus on Suitable Roles – Prioritize administrative, customer service, and lower-skilled jobs that can be handled part-time without impacting productivity.

-   Flexible Scheduling – Attract strong candidates by offering adaptable work hours and relevant benefits.

-   Track Impact – Regularly assess productivity and cost savings to ensure the policy remains effective.
